Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-05-08 paper/ms.tex
update ms

Exoplanet science is in an age of abundance, fueled by the discovery
of thousands of worlds orbiting close to their host stars
\citep{Borucki10,2015JATIS...1a4003R}.  Most of these exoplanets are
billions of years old, similar to the Earth.  This fact leaves many
gaps in our knowledge of what exact dynamical and physical processes
produce these objects.    The reason is that processes such as thermal
cooling \citep{2007ApJ...659.1661F}, atmospheric loss
\citep{2019AREPS..47...67O}, giant impacts
\citep{2014prpl.conf..595R}, and dynamical instabilities
\citep{2017MNRAS.470.1750I} are expected to be most efficient over
$\approx$10\,Myr to $\approx$1\,Gyr timescales; when we observe most
known exoplanets, most of the action is in the past.

Finding and characterizing young ($<$1\,Gyr) exoplanets is one
potential path toward building a timeline for exoplanet evolution.
Informative individual young systems include V1298~Tau, an adolescent
resonant chain of four transiting puffy planets that is a likely
precusor to the compact multiplanet systems \citep{David_2019}, and
HIP~67522, a Jupiter-sized planet with a Neptunian mass
(\citealt{Rizzuto_2020}; Thao et al.~submitted).  Population-level
analyses have similarly suggested differences in the size distribution
of young exoplanets relative to their older counterparts
\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.

@@

To begin, imagine that the ages of nearby stars in the Galaxy are
uniformly distributed between 0 and 10\,Gyr
\citep[e.g.~Fig~17 of][]{Nordstrom_2004}.  This approximation clarifies the
rarity of young stars: under these assumptions, $\approx$1\% of nearby
stars have $t$$<$100\,Myr, and $\approx$10\% have $t$$<$1\,Gyr.
Studies of currently forming protoplanets \citep{2018A&A...617A..44K},
or of exoplanets evolving just after disk dispersal
\citep[e.g.][]{2022MNRAS.512.5067K}, are thereby capped in their
maximum achievable sample sizes by the tyranny of the Galactic star
formation rate.

Despite the observational challenges, young close-in planet discovery
has matured as a field over the past decade, primarily due to
Kepler, K2, TESS, and Gaia
\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Barber_2022,Bouma_2022a,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
The strategy pursued by most groups during the 2010s was to focus on
stars with known ages---obvious members of open clusters---and to
search them for transiting planets.  The resulting stellar (and assumed planetary) ages are
precise at the $\approx$10\% level.  A
more recent development, facilitated by the second Gaia data release,
is the idea that finding transiting planets can often be easier than
finding and characterizing the host stellar association
\citep[e.g.][]{Tofflemire_2021}.

The main problem with the ``cluster-first'' approach is that most
nearby young stars are in the field.  Only $\approx$1\% of
stars within the nearest $\approx$500\,pc have been associated
with their birth cluster
\citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
One can appreciate the problem by querying NASA's Exoplanet Archive
\citep[NEA;][]{2013PASP..125..989A} for transiting planets younger
than 1\,Gyr.  Imposing reasonable precision requirements
(e.g.~$t$$<$1\,Gyr at 2$\sigma$, or else ages precise to say
S/N$=$3) gives $\approx$50 ``young'' transiting planets at the time
of writing.  Most of these planets are in clusters, and were
discovered by Kepler, K2, or TESS.  However, combined, those surveys
have discovered $\approx$5{,}000 planets.  Unless there is a major
bias against the discovery of sub-gigayear old planets with precisely
constrained ages, a constant star formation history implies that
an order of magnitude more young
transiting planets should be known.

This study aims to resolve two main questions.  First, how wrong it is
to assume a uniform age distribution for stars in the galactic thin
disk, particularly young stars?  Second, where are the missing young
transiting planets?  Briefly, we will find that uniform is incorrect
by a factor of a few, and that the bias against the discovery of young
transiting planets is, primarily, in the measurement of
theirs ages. %, rather than in the sensitivity of the transit searches.

Given that Kepler's main mission ended in 2013, one might be
forgiven for assuming that the ages of the Kepler planets are already
known.  Of course, they have been studied: \citet{Berger_2020b_rpage}
and \citet{Petigura_2022} for instance calculated isochrone ages for
Kepler Objects of Interest (KOIs).  These ages are most precise for
stars whose luminosities and temperatures separate them from the
main-sequence.  \citet{David_2021} leveraged both isochronal ages from
an earlier analogous study \citep{Fulton_2018}, and
used a relative gyrochrone age-ranking approach that sorted stars
based on whether they rotated faster or slower than a gyrochrone age
of $\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.  These studies
were conducted in the broader context of investigating trends in
exoplanet demographics, which also depend on stellar metallicity and
mass \citep[e.g.][]{Petigura_2018}.  These studies offered limited
sensitivity to the ages of the youngest planets, due to the nature of
isochronal dating.

Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
between 1-4\,Gyr, and provides more complex asymmetric posteriors at earlier
times \citep{Bouma_2023}.  The idea of using a star's spin-down as a
clock has a rich history
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
and physics-based models for the spin-down itself can clarify many
aspects of how the stellar winds, internal structure, and magnetic
dynamo all evolve
\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  In the
context of Kepler, empirical models have been previously applied to
derive rotation-based ages for subsets of the Kepler data
\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.  ``Gyro-kinematic'' ages
have also been reported \citep{2021AJ....161..189L}.
There have been
major improvements since these pioneering studies in:
\begin{enumerate}[label={\it \roman*)},leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The extent of the Kepler Object of Interest catalog, and our vetting of false
    positives within it;
  %
  \item Available measurements of stellar rotation periods
    in open clusters
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
  %
  \item Our ability to model not only the mean evolution but also the
    intrinsic dispersion in rotation through time
    \citep{Bouma_2023}.  For Sun-like stars, the new model is broadly
    consistent with older calibrations \citep[e.g.][]{Mamajek_2008}.
    For late G dwarfs and K dwarfs, the age calibration
    differs by factors of a few;
  %
  \item Our ability to identify binary stars, which can be problematic
    for estimates of rotation-based ages.
\end{enumerate}
Today, FGK star rotation-based ages are
empirically tied to the cluster age scale, and
are limited in precision by the intrisic scatter in the $P_{\rm rot}$
sequences, and in accuracy by systematic uncertainties in the evolving
spin-down rates, and in the cluster ages themselves.

Beyond improvements in rotation-based ages, recent years have also
yielded improvements in the lithium age scale.
Lithium ages include two qualitatively distinct regimes: depletion
boundary ages derived for M-dwarfs in star clusters, and the less
precise decline-based ages for individual field FGK stars
\citep{Soderblom_2010}.  Here we focus on the latter approach, which
relies on the observation that the Li abundances of
partially-convective stars gradually decline as they age
\citep[e.g.][]{2005A&A...442..615S}.  The theoretical explanation for
this decline is debated
\citep[e.g.][]{1995ApJ...441..865C,2010ApJ...716.1269D,2019MNRAS.485.4052C}
Empirical understanding has however recently improved due to the work
by \citet{Jeffries_2023}, who built a model for how the equivalent
width (EW) of the \ion{Li}{1} 6708\,\AA\ doublet evolves as a function
of stellar effective temperature and age for a set of 6{,}200 stars in
52 open clusters.  Two-sided lithium ages are useful for Kepler (FGK)
stars between $\approx$0.03-0.5\,Gyr, with a strong dependence on
spectral type since the K-dwarfs lose their surface lithium much
faster than G-dwarfs.  The precision of the ages reported by the
\citet{Jeffries_2023} method in this regime are in the range of
0.3-0.5~dex.

We discuss our methods for selecting stellar and planetary samples in
Section~\ref{sec:selection}, and describe
the origin of our adopted stellar parameters other than ages in
Section~\ref{sec:stellarprops}.
We describe and validate our age-dating methods in
Sections~\ref{sec:rotage} and~\ref{sec:liage}.
The population-level trends for both the parent stellar sample and
the planet sample are discussed in Section~\ref{sec:results}.
A few conclusions are offered in Section~\ref{sec:conclusions}.

@@

\begin{figure*}[!t]
  \begin{center}
    \subfloat{
      \includegraphics[width=0.49\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
    }
    \subfloat{
    \includegraphics[width=0.49\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
    }
  \end{center}
  \vspace{-0.5cm}
  \caption{
    {\bf The stars.}  Our analysis focuses on stars observed by Kepler
    with previously reported rotation periods (dark gray points).
    The rotation periods were primarily measured by \citet{Santos_2019} and \citet{Santos_2021}.
    About half of these rotators are suitable for gyrochronology
    (black points), based on factors including their temperatures,
    surface gravities, non-binarity, and proximity to the main
    sequence (see Section~\ref{subsec:flags}).   Some are
    ``confirmed'' or ``candidate'' Kepler Objects of Interest that
    meet an additional set of planetary quality criteria (blue points;
    Section~\ref{subsec:plflags}).  Surface gravities and effective
    temperatures were derived photometrically by \citet{Berger_2020a_catalog}.
  }
  \label{fig:stellarprops}
\end{figure*}

\begin{figure*}[!t]
  \begin{center}
    \leavevmode
    \subfloat{
      \includegraphics[width=0.49\textwidth]{gyromodeldispersion.pdf}
    }
    \subfloat{
      \includegraphics[width=0.49\textwidth]{li_vs_teff_koi_X_S19S21dquality_eagles_showpoints_nodata.pdf}
    }
  \end{center}
  \vspace{-0.6cm}
  \caption{
    {\bf The models.}
    Points represent $10^4$ draws from model probability distributions
    that have been fitted to rotation periods
    \citep{Bouma_2023} and lithium equivalent widths
    \citep[EWs;][]{Jeffries_2023} of stars in open clusters.  Lines
    are the ``mean models'' at various ages; the asymmetric intrinsic
    dispersion in rotation and lithium about these mean models, which
    is what the models fit, sets the theoretical precision floor for
    both age-dating methods.    Additional sources of uncertainty,
    including measurement uncertainty, impose further limits on
    achievable age precision.  These models are calibrated using open
    clusters younger than 4\,Gyr.  The displayed points assume a
    uniform distribution in temperature for visual clarity.  The sizes
    of the points are the same in each panel, so that low apparent
    density signifies greater dispersion around the mean.
    \label{fig:models}
  }
\end{figure*}

\begin{figure*}[!t]
  \begin{center}
    \subfloat{
      \includegraphics[width=0.48\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
    }
    \subfloat{
      \includegraphics[width=0.48\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
    }
  \end{center}
  \vspace{-0.5cm}
  \caption{
    {\bf Rotation periods for Kepler target stars (left) and known
    planet hosts (right)}.  As in Figure~\ref{fig:stellarprops}, periods are from \citet{Santos_2019}
    and \citet{Santos_2021} (\citetalias{Santos_2019} and
    \citetalias{Santos_2021}).  The gray lines are ``mean fits'' 
    to the slow rotation sequences of M67,
    Ruprecht~147, NGC~6811, Praesepe, and a trio of 120\,Myr open
    clusters.  The stellar sample includes only apparently single stars
    near the main sequence with $\log g$$>$4.2, RUWE$<$1.4,
    and 3800$<$$T_{\rm eff}$/K$<$6200.  The displayed planets require
    the same stellar cuts, and include only the confirmed and candidate planets
    described in Section~\ref{subsec:plflags}.
    An order of magnitude estimate for the number of young planets discovered
    by Kepler follows by counting planets below the mean rotational isochrones.
  }
  \label{fig:prot_vs_teff}
\end{figure*}

\begin{figure}[!t]
  \begin{center}
    \leavevmode
    \includegraphics[width=0.48\textwidth]{li_vs_teff_koi_X_JUMP_eagles_logy.pdf}
  \end{center}
  \vspace{-0.6cm}
  \caption{
    {\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
      for planet-hosting stars.} These measurements were made from
    Keck/HIRES spectra collected between 2009 and 2024 that comprised
    \nhireshours\,hours of open-shutter time.  Lines are the ``mean''
    isochrones from the same models as in Figure~\ref{fig:models}; the
    intrinsic dispersion beyond $\approx$1\,Gyr becomes much larger
    than changes in the mean.  The logarithmic scale highlights the
    dynamic range of the detections and upper limits.  Some stars with
    lithium detections do not have detected rotation periods, and
    vice-versa.
    \label{fig:li_vs_teff}
  }
\end{figure}

This work is focused on Kepler stars for which ages can be inferred
using either rotation, lithium, or both.  These ``age-dateable
stars'' are a minority of the \nkeplerstars\ Kepler targets.  Rotation
periods have been reported for roughly one in three Kepler targets
\citep[e.g.][]{McQuillan_2014,Santos_2021}.  High-resolution spectra,
which can yield measurements of the \ion{Li}{1} 6708\,\AA\ doublet, 
have only been acquired for the Kepler objects of interest
(KOIs), which comprise a few percent of Kepler's targets.
In the following, we will describe the set of stars that we
adopt with measured rotation periods (Section~\ref{subsec:rotsel}),
the set of objects we adopt as planets
(Section~\ref{subsec:planetsel}), and the set of planets with
high-resolution spectra suitable for lithium analysis
(Section~\ref{subsec:lithiumsel}).

@@

To select stars with rotation periods, we turn to previous work.
Many investigators have derived rotation periods both specifically of
the Kepler Objects of Interest
\citep{McQuillan_2013,Walkowicz_2013,Mazeh_2015,Angus_2018,David_2021},
and also of the broader set of all Kepler target stars
\citep{McQuillan_2014,Reinhold_2015,Santos_2019,Santos_2021}.
These studies used a range of detection methods and selection
functions.
We are interested in understanding the age distribution of not only
the KOIs but also the Kepler target stars.
Out of these studies, the only homogeneous analysis of both Kepler
targets and KOIs were the studies by
\citet{Santos_2019} (\citetalias{Santos_2019}),
and \citet{Santos_2021} (\citetalias{Santos_2021}).
\citetalias{Santos_2019} and
\citetalias{Santos_2021} combined a wavelet analysis and
autocorrelation-based approach, and cumulatively reported rotation
periods for 55{,}232 main-sequence and subgiant FGKM stars.
They included known KOIs and binaries in their analysis, and assigned
them specific quality flags, while removing transits and eclipses
during the stellar rotation measurement process.

@@

We therefore adopt the results of \citetalias{Santos_2019} and
\citetalias{Santos_2021} as our default rotation periods.
\citetalias{Santos_2021} provided a comparison against
\citetalias{McQuillan_2014}; the brief summary is that the periods
agree for 99.0\% of the 31{,}038 period detections in common between
the two studies.  \citetalias{Santos_2021} classified the 2{,}992
remaining stars from \citetalias{McQuillan_2014} as not showing
rotation periods based on updated knowledge of contaminants
(e.g.~giant stars and eclipsing binaries) and visual inspection.  In
addition, \citetalias{Santos_2021} reported rotation periods for
24{,}182 main-sequence and subgiant FGKM stars that were not reported
as periodic by \citetalias{McQuillan_2014}.  Many of these reported
detections have lower variability amplitudes and longer periods than
those reported by \citetalias{McQuillan_2014}.  A relevant flag for
stars missing \citetalias{McQuillan_2014} periods is included in
Table~\ref{tab:stars}.

Analyzing the compilation of \citetalias{Santos_2019} and
\citetalias{Santos_2021} rotation periods for the KOI hosts, we
noticed that some known KOIs with rotation periods were missing.
This is not surprising, in that the rotation periods of KOIs have
received considerably more scrutiny than those of ordinary Kepler
stars.
We therefore decided to split our subsequent analysis into a
homogeneous portion that used only the \citetalias{Santos_2019} and
\citetalias{Santos_2021} data, and an inhomogeneous portion that also
considered a broader set of available KOI rotation periods.
For the latter portion, we first included 32 KOIs with orbital and
rotation periods within $\approx$20\% that had been
excluded %\footnote{Excluding stars in this category would impose a bias
%against young close-in planets, due to the natural commensurability
%between the rotation rates of young stars, and Kepler's increased
%sensitivity to short-period objects.  Although some of these KOIs are
%known false positives, our analysis excludes such objects at later
%steps.}%  As one motivating example, this subset of stars includes {\it
%e.g.}~Kepler~1643 (KIC~8653134), with $P_{\rm orb}$=5.34\,days and
%$P_{\rm rot}$=5.05\,days, which is known to be $\approx$40\,Myr old
%based on membership in RSG-5 \citep{Bouma_2022b}.  }
from the
\citetalias{Santos_2019} and \citetalias{Santos_2021} catalogs
(A.~Santos, private communication).
We then incorporated an additional \nnewdavidtwentyone\ rotation
periods for KOI hosts that
\citet{David_2021} had described as either ``reliable'' or ``highly
reliable'' in their visual analysis of previously reported KOI
rotation periods from \citet{McQuillan_2013,Walkowicz_2013,Mazeh_2015}
and \citet{Angus_2018}.
Inclusion of these additional KOI rotation periods is a
supplementary measure aimed at completeness in our final KOI age
catalog; the provenances of the individual adopted periods are all
noted in the relevant tables.

Finally, since our scope is focused on rotation-based age dating, 
we restricted our attention to stars with reported $P_{\rm
rot}$$<$45\,days.
FGK stars with slower rotation periods spin much slower than the stars
in the oldest open clusters currently used to calibrate empirical
gyrochronal models.

The resulting tally of stars from
\citetalias{Santos_2019}, \citetalias{Santos_2021}, and our corrected
KOI list is \nuniqstarsantosrot\ Kepler stars with rotation periods.

>>>>

Exoplanet science is in an age of abundance, fueled by the discovery
of thousands of worlds orbiting close to their host stars
\citep{Borucki10,2015JATIS...1a4003R}.  Most of these exoplanets are
billions of years old.  This fact leaves many gaps in our knowledge of
the exact physical and dynamical pathways that produce these objects.
The reason is that processes such as thermal cooling
\citep{2007ApJ...659.1661F}, atmospheric loss
\citep{2019AREPS..47...67O}, giant impacts
\citep{2014prpl.conf..595R}, and dynamical instabilities
\citep{2017MNRAS.470.1750I} are expected to be most efficient over
$\approx$10\,Myr to $\approx$1\,Gyr timescales; for most known
exoplanets, most of the action has already happened.

Finding and characterizing young ($<$1\,Gyr) exoplanets can therefore
help in building the timeline for exoplanet evolution.  Informative
individual young systems include V1298~Tau, an adolescent resonant
chain of four transiting puffy planets that is a likely precusor to
the compact multiplanet systems \citep{David_2019}, and HIP~67522, a
Jupiter-sized planet with a Neptunian mass (\citealt{Rizzuto_2020};
Thao et al.~submitted).  Population-level analyses have similarly
suggested differences in the size distribution of young exoplanets
relative to their older counterparts
\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.

@@

To begin, imagine that the ages of nearby stars in the Galaxy are
uniformly distributed between 0 and 10\,Gyr \citep[e.g.~Fig~17
of][]{Nordstrom_2004}.  This approximation clarifies the rarity of
young stars: under this assumption, $\approx$1\% of nearby stars have
$t$$<$100\,Myr, and $\approx$10\% have $t$$<$1\,Gyr.  Studies of
currently forming protoplanets \citep{2018A&A...617A..44K}, or of
exoplanets evolving just after disk dispersal
\citep[e.g.][]{2022MNRAS.512.5067K}, are thereby capped in their
maximum achievable sample sizes by the tyranny of the Galactic star
formation rate.

Despite the observational challenges, young close-in planet discovery
has matured over the past decade, primarily due to Kepler, K2, TESS,
and Gaia
\citep[e.g.][]{Meibom_2013,Mann_K2_25_2016,Mann_2017,Curtis_2018,Livingston_2018,David_2019,Bouma_2020_toi837,Rizzuto_2020,Plavchan_2020,Newton_2021,Nardiello_2022,Tofflemire_2021,Barber_2022,Bouma_2022a,Bouma_2022b,Zhou_2022,Zakhozhay_2022,Wood_2023}.
The strategy pursued by most groups during the 2010s was to focus on
stars with known ages---obvious members of open clusters---and to
search them for transiting planets.  The resulting stellar (and
assumed planetary) ages are precise at the $\approx$10\% level.  A
more recent development, facilitated by the second Gaia data release,
is the idea that finding transiting planets can often be easier than
finding and characterizing the host stellar association
\citep[e.g.][]{Tofflemire_2021}.

The main problem with the ``cluster-first'' approach is that most
nearby young stars are in the field.  Only $\approx$1\% of stars
within the nearest $\approx$500\,pc are associated with their birth
cluster
\citep[e.g.][]{Zari_2018,CantatGaudin_2020,Kounkel_2020,Kerr_2021}.
One can appreciate the problem by querying NASA's Exoplanet Archive
\citep[NEA;][]{2013PASP..125..989A} for transiting planets younger
than 1\,Gyr.  Imposing reasonable precision requirements
(e.g.~$t$$<$1\,Gyr at 2$\sigma$, or else ages precise to say S/N$=$3)
gives $\approx$50 ``young'' transiting planets at the time of writing.
Most of these planets are in clusters, and were found by Kepler, K2,
or TESS.  However, combined, those surveys have discovered
$\approx$5{,}000 planets.  Unless there is a major bias against the
discovery of sub-gigayear old planets with precisely constrained ages,
a constant star formation history implies that an order of magnitude
more young transiting planets should be known.

This study aims to resolve two main questions.  First, how wrong it is
to assume a uniform age distribution for stars in the galactic thin
disk?  Second, where are the missing young transiting planets?
Briefly, we will find that uniform is incorrect by a factor of a few,
and that at least one bias against the discovery of young transiting
planets is in the measurement of their ages.

Given that Kepler's main mission ended in 2013, one might be
forgiven for assuming that the ages of the Kepler planets are already
known.  Of course, they have been studied: \citet{Berger_2020b_rpage}
and \citet{Petigura_2022} for instance calculated isochrone ages for
Kepler Objects of Interest (KOIs).  These ages are most precise for
stars whose luminosities and temperatures separate them from the
main-sequence.  \citet{David_2021} leveraged isochronal ages from
an earlier analogous study \citep{Fulton_2018}, and also
used a relative gyrochrone age-ranking approach.
%that sorted stars
%based on whether they rotated faster or slower than a gyrochrone age
%of $\approx$2.6\,Gyr \citep{Meibom_2015,Curtis_2020}.
These studies were conducted in the broader context of investigating
trends in exoplanet demographics, which also depend on stellar
metallicity and mass \citep[e.g.][]{Petigura_2018}.  These studies
offered limited sensitivity to the ages of $\lesssim$1\,Gyr planets,
due to the nature of isochronal dating.

Gyrochronology offers ages precise to $\lesssim$30\% for FGK stars
between 1-4\,Gyr, and provides more complex asymmetric posteriors at earlier
times \citep{Bouma_2023}.  The idea of using a star's spin-down as a
clock has a rich history
\citep{Skumanich_1972,Noyes_1984,Kawaler_1989,Barnes03,Mamajek_2008,Angus_2015},
and physics-based models for the spin-down itself can clarify many
aspects of how the stellar winds, internal structure, and magnetic
dynamo all evolve
\citep[e.g.][]{Matt_2015,Gallet_Bouvier_2015,Spada_2020}.  In the
context of Kepler, empirical models have been previously applied to
derive rotation-based ages for subsets of the Kepler data
\citep{Walkowicz_2013,Reinhold_2015,Angus_2018}.  ``Gyro-kinematic'' ages
have also been reported \citep{2021AJ....161..189L}.
There have been
major improvements since these pioneering studies in:
\begin{enumerate}[label={\it \roman*)},leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The extent of the Kepler Object of Interest catalog, and our vetting of false
    positives within it;
  %
  \item Available measurements of stellar rotation periods
    in open clusters
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B};
  %
  \item Our ability to model not only the mean evolution but also the
    intrinsic dispersion in rotation through time
    \citep{Bouma_2023}.  For Sun-like stars, the new model is broadly
    consistent with older calibrations \citep[e.g.][]{Mamajek_2008}.
    For late G dwarfs and K dwarfs, the new calibration
    differs by factors of a few;
  %
  \item Our ability to identify binary stars, which can be problematic
    for estimates of rotation-based ages.
\end{enumerate}
These improvements form the basis for why we believe the present
analysis to be necessary.
%Today, FGK star rotation-based ages are
%empirically tied to the cluster age scale, and
%are limited in precision by the intrisic scatter in the $P_{\rm rot}$
%sequences, and in accuracy by systematic uncertainties in the evolving
%spin-down rates, and in the cluster ages themselves.

However, beyond improvements in rotation-based ages, recent years have
also yielded improvements in the lithium age scale.  Lithium ages
include two qualitatively distinct regimes: depletion boundary ages
for M-dwarfs in star clusters, and the less precise decline-based ages
for individual field FGK stars \citep{Soderblom_2010}.  Here we focus
on the latter approach, which relies on the observation that the Li
abundances of partially-convective stars gradually decline as they age
\citep[e.g.][]{2005A&A...442..615S}.  The theoretical explanation for
this decline is debated
\citep[e.g.][]{1995ApJ...441..865C,2010ApJ...716.1269D,2019MNRAS.485.4052C}
Empirical understanding has however improved due to the work by
\citet{Jeffries_2023}, who built a model for how the equivalent width
(EW) of the \ion{Li}{1} 6708\,\AA\ doublet evolves as a function of
stellar effective temperature and age for a set of 6{,}200 stars in 52
open clusters.  Two-sided lithium ages are useful for Kepler (FGK)
stars between $\approx$0.03-0.5\,Gyr, with a strong dependence on
spectral type since the K-dwarfs lose their surface lithium much
faster than G-dwarfs.  The precision of the ages reported by the
\citet{Jeffries_2023} method in this regime are in the range of
0.3-0.5~dex.

We discuss our methods for selecting the star and planet samples in
Section~\ref{sec:selection}, and describe the origin of our adopted
stellar parameters other than ages in Section~\ref{sec:stellarprops}.
We describe and validate our age-dating methods in
Sections~\ref{sec:rotage} and~\ref{sec:liage}.  The population-level
trends for both stars and planets are described in
Sections~\ref{sec:results} and~\ref{sec:disc}, and a few conclusions
are offered in Section~\ref{sec:conclusions}.

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.49\textwidth]{st_params_M_G_vs_dr3_bp_rp.pdf}
		}
		\subfloat{
			\includegraphics[width=0.49\textwidth]{st_params_adopted_logg_vs_adopted_Teff.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
		{\bf The stars.}  Our analysis focuses on stars observed by Kepler
		with previously reported rotation periods (dark gray points).
		The rotation periods were primarily measured by \citet{Santos_2019,Santos_2021}.
		About half of these rotators are suitable for gyrochronology
		(black points), based on factors including their temperatures,
		surface gravities, non-binarity, and proximity to the main
		sequence (see Section~\ref{subsec:flags}).   Some are
		``confirmed'' or ``candidate'' Kepler Objects of Interest that
		meet an additional set of planetary quality criteria (blue points;
		Section~\ref{subsec:plflags}).  Surface gravities and effective
		temperatures were derived photometrically by \citet{Berger_2020a_catalog}.
	}
	\label{fig:stellarprops}
\end{figure*}

This work is focused on Kepler stars for which ages can be inferred
using either rotation, lithium, or both.  These ``age-dateable
stars'' are a minority of the \nkeplerstars\ Kepler targets.  Rotation
periods have been reported for $\approx$one in three Kepler targets
\citep[e.g.][]{McQuillan_2014,Santos_2021}.  High-resolution spectra
suitable for measurement of the \ion{Li}{1} 6708\,\AA\ doublet
have only been acquired for the Kepler objects of interest
(KOIs), which comprise a few percent of Kepler's targets.
In the following, we will describe the set of stars that we
adopt with measured rotation periods (Section~\ref{subsec:rotsel}),
the set of objects we adopt as planets
(Section~\ref{subsec:planetsel}), and the set of planets with
high-resolution spectra suitable for lithium analysis
(Section~\ref{subsec:lithiumsel}).
Figure~\ref{fig:stellarprops} provides a visualization of the various samples.

@@

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.48\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
		}
		\subfloat{
			\includegraphics[width=0.48\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
		{\bf Rotation periods for Kepler target stars (left) and known
			planet hosts (right)}.  
		The gray lines are ``mean fits'' 
		to the slow rotation sequences of M67,
		Ruprecht~147, NGC~6811, Praesepe, and a trio of 120\,Myr open
		clusters.  The stellar sample includes only apparently single stars
		near the main sequence with $\log g$$>$4.2, RUWE$<$1.4,
		and 3800$<$$T_{\rm eff}$/K$<$6200.  The displayed planet-hosts require
		the same stellar cuts, and include only the confirmed and candidate planets
		described in Section~\ref{subsec:plflags}.
		An order of magnitude estimate for the number of young planets discovered
		by Kepler follows by counting planets below the mean rotational isochrones.
	}
	\label{fig:prot_vs_teff}
\end{figure*}

To select stars with rotation periods, we turn to previous work.
Many investigators have derived rotation periods both specifically for
KOIs
\citep{McQuillan_2013,Walkowicz_2013,Mazeh_2015,Angus_2018,David_2021},
and also for the broader set of all Kepler target stars
\citep{McQuillan_2014,Reinhold_2015,Santos_2019,Santos_2021}.
These studies used a range of detection methods and selection
functions.
We are interested in understanding the age distribution of all Kepler
target stars irrespective of KOI status.
Out of these studies, the only homogeneous analysis of both Kepler
targets and KOIs were the studies by \citet{Santos_2019,Santos_2021}
(\citetalias{Santos_2019} and \citetalias{Santos_2021}).
\citetalias{Santos_2019} and
\citetalias{Santos_2021} combined a wavelet analysis and
autocorrelation-based approach, and cumulatively reported rotation
periods for 55{,}232 main-sequence and subgiant FGKM stars.
They included known KOIs and binaries in their analysis, and assigned
them specific quality flags, while removing transits and eclipses
during the stellar rotation measurement process.

@@

We therefore adopt the results of \citetalias{Santos_2019} and
\citetalias{Santos_2021} as our default rotation periods.
\citetalias{Santos_2021} provided a comparison against
\citetalias{McQuillan_2014}; the brief summary is that the periods
agree for 99.0\% of the 31{,}038 period detections in common between
the two studies.  \citetalias{Santos_2021} classified the 2{,}992
remaining stars from \citetalias{McQuillan_2014} as not showing
rotation periods based on updated knowledge of contaminants
(e.g.~giant stars and eclipsing binaries) and visual inspection.  In
addition, \citetalias{Santos_2021} reported rotation periods for
24{,}182 main-sequence and subgiant FGKM stars that were not reported
as periodic by \citetalias{McQuillan_2014}.  Many of these reported
detections have lower variability amplitudes and longer periods than
those reported by \citetalias{McQuillan_2014}. 
%A relevant flag for
%stars missing \citetalias{McQuillan_2014} periods is included in
%Table~\ref{tab:stars}.

Analyzing the compilation of \citetalias{Santos_2019} and
\citetalias{Santos_2021} rotation periods for the KOI hosts, we
noticed that some known KOIs with rotation periods were missing.
This is not surprising, in that the rotation periods of KOIs have
received considerably more scrutiny than those of ordinary Kepler
stars.
We therefore decided to split our subsequent analysis into a
homogeneous portion that used only the \citetalias{Santos_2019} and
\citetalias{Santos_2021} data, and an inhomogeneous portion that also
considered a broader set of available KOI rotation periods.
For the latter portion, we first included 32 KOIs with orbital and
rotation periods within $\approx$20\% that had been
excluded %\footnote{Excluding stars in this category would impose a bias
%against young close-in planets, due to the natural commensurability
%between the rotation rates of young stars, and Kepler's increased
%sensitivity to short-period objects.  Although some of these KOIs are
%known false positives, our analysis excludes such objects at later
%steps.}%  As one motivating example, this subset of stars includes {\it
%e.g.}~Kepler~1643 (KIC~8653134), with $P_{\rm orb}$=5.34\,days and
%$P_{\rm rot}$=5.05\,days, which is known to be $\approx$40\,Myr old
%based on membership in RSG-5 \citep{Bouma_2022b}.  }
from the
\citetalias{Santos_2019} and \citetalias{Santos_2021} catalogs
(A.~Santos, private communication).
We then incorporated an additional \nnewdavidtwentyone\ rotation
periods for KOI hosts that
\citet{David_2021} had described as either ``reliable'' or ``highly
reliable'' in their visual analysis of previously reported KOI
rotation periods from \citet{McQuillan_2013,Walkowicz_2013,Mazeh_2015}
and \citet{Angus_2018}.
Inclusion of these additional KOI rotation periods is a
supplementary measure aimed at completeness in our final KOI age
catalog; the provenances of the individual adopted periods are 
noted in the relevant tables.

Finally, since our scope is focused on rotation-based age dating, 
we restricted our attention to stars with reported $P_{\rm
rot}$$<$45\,days.
FGK stars with slower rotation periods spin much slower than the stars
in the oldest open clusters currently used to calibrate our 
gyrochronology model.
Figure~\ref{fig:prot_vs_teff} shows the resulting
\nuniqstarsantosrot\ Kepler stars with rotation periods 
compiled from
\citetalias{Santos_2019}, \citetalias{Santos_2021}, and our extended
KOI list.
